---
title: "MA615 MIDTERM PROJECT"
author: "TSZ WAI NG"
format: html
editor: visual
---

## Assignment

## Data Cleaning

```{r}
#| label: Load libraries
#| warning: false
#| message: false
#| echo: false

library(knitr)  
library(kableExtra)
library(tidyverse)
library(stringr)
library(readr)
library(ggplot2)
library(forecast)
library(dplyr)
```

<!-- Read the file -->

```{r warnming=FALSE, message=FALSE}
#FEMA
DisasterDeclarations <- read_csv("/Users/julie/Desktop/MA615/MID-TERM PROJECT/FEMA DATASETS/DisasterDeclarationsSummaries.csv")
FEMAWeb <- read_csv("/Users/julie/Desktop/MA615/MID-TERM PROJECT/FEMA DATASETS/FemaWebDisasterSummaries.csv")

#StormEventsDetail
StormDetails2020 <- read_csv("/Users/julie/Desktop/MA615/MID-TERM PROJECT/NOAA STORM EVENTS DATABASE/StormEvents_details-ftp_v1.0_d2020_c20230927.csv")
StormDetails2021 <- read_csv("/Users/julie/Desktop/MA615/MID-TERM PROJECT/NOAA STORM EVENTS DATABASE/StormEvents_details-ftp_v1.0_d2021_c20231017.csv")

#StormEventsLocation
StormLocation2020 <- read_csv("/Users/julie/Desktop/MA615/MID-TERM PROJECT/NOAA STORM EVENTS DATABASE/StormEvents_locations-ftp_v1.0_d2020_c20230927.csv")
StormLocation2021 <- read_csv("/Users/julie/Desktop/MA615/MID-TERM PROJECT/NOAA STORM EVENTS DATABASE/StormEvents_locations-ftp_v1.0_d2021_c20231017.csv")

#StormEventsFatalities
StormFatalities2020 <- read_csv("/Users/julie/Desktop/MA615/MID-TERM PROJECT/NOAA STORM EVENTS DATABASE/StormEvents_fatalities-ftp_v1.0_d2020_c20230927.csv")
StormFatalities2021 <- read_csv("/Users/julie/Desktop/MA615/MID-TERM PROJECT/NOAA STORM EVENTS DATABASE/StormEvents_fatalities-ftp_v1.0_d2021_c20231017.csv")
```

I only interested in flood between 2020 and 2021, so I will do some data cleaning to keep the information that are related

```{r warning=FALSE, message=FALSE}
#StormEventsDetail
StormDetails2020 <- StormDetails2020 %>% filter(EVENT_TYPE == "Flash Flood")
StormDetails2021 <- StormDetails2021 |> filter(EVENT_TYPE == "Flash Flood")

#FEMA
DisasterDeclarations <- DisasterDeclarations %>% filter(incidentType == "Flood" & year(incidentBeginDate) %in% c(2020,2021))
```

Drop all the columns that only contains NA value

```{r warning=FALSE, message=FALSE}
#| label: drop one-item columns
#| echo: false

## define function
drop_null_col <- function(df) {
  col_name <- NULL
  for (i in 1:ncol(df)) {
    if (all(is.na(df[, i]))) {
      col_name <- c(col_name, colnames(df)[i])
    }
  }
if(is.null(col_name)){return("No Columns to drop")}else{
   drp = data.frame(col_name)
   return(drp)
   }
}

#Ddrop columns that only have NA for StromEventsDetail
SD2020 <- drop_null_col(StormDetails2020)
SD2020 <- SD2020$col_name
StormDetails2020 <- StormDetails2020 |> select(!all_of(SD2020))

SD2021 <- drop_null_col(StormDetails2021)
SD2021 <- SD2021$col_name
StormDetails2021 <- StormDetails2021 |> select(!all_of(SD2021))

#Ddrop columns that only have NA for data from FEMA
DD <- drop_null_col(DisasterDeclarations)
DD <- DD$col_name
DisasterDeclarations <- DisasterDeclarations |> select(!all_of(DD))
```

## Initial questions

1.  I am interested in how long the flooding lasts in every state.

```{r warning=FALSE, message=FALSE}
#FEMA data
FEMABegainDate <- as.Date(DisasterDeclarations$incidentBeginDate)
FEMAEndDate <- as.Date(DisasterDeclarations$incidentEndDate)
Duration <- FEMAEndDate - FEMABegainDate
DisasterDeclarations <- cbind(DisasterDeclarations,Duration)
DisasterDeclarations$Duration <- as.numeric(gsub("[^0-9]", "",DisasterDeclarations$Duration))

DD_count <- DisasterDeclarations %>%
  group_by(incidentBeginDate,state) %>%
  count(duration = Duration)
```

```{r}
ggplot(DD_count) +
  aes(x = state, y = duration, fill = state) +
  geom_col() +
  scale_fill_hue(direction = 1) +
  theme_classic() + labs(title="Flooding Duration in each State")

ggplot(DisasterDeclarations) +
  aes(x = incidentBeginDate,y=Duration) +
  geom_point() +
  theme_classic()+ labs(title = "Flooding declaration date")
```

The first graph shows that WA state has the most flooding duration, PR and TN have the lowest day of flooding. The reasons that WA has a long time flooding duration is because it has lots of rivers. The second plot indicates that the flooding usually has a higher duration and more flooding during first half of the year

```{r warning=FALSE, message=FALSE}
filterfema <- FEMAWeb[FEMAWeb$disasterNumber %in% DisasterDeclarations$disasterNumber,]

filterfema$state <- DisasterDeclarations$state[match(filterfema$disasterNumber, DisasterDeclarations$disasterNumber)]

ggplot(filterfema) +
  aes(
    x = state,
    y = totalObligatedAmountCatC2g,
    colour = state
  ) +
  geom_point(shape = "circle", size = 3) +
  scale_color_hue(direction = 1) +
  labs(
    title = "How much each state needs to rebuild public infrastructure"
  ) +
  theme_classic()

ggplot(filterfema) +
  aes(
    x = state,
    y = totalObligatedAmountCatAb,
    colour = state
  ) +
  geom_point(shape = "circle", size = 3) +
  scale_color_hue(direction = 1) +
  labs(
    title = "How much each state needs for EMERGENCY WORK"
  ) +
  theme_classic()

```

From this graph, it shows that KY needs the most
